MassWateR: Improving Quality Control, Analysis, and Sharing of Water Quality Data

### 2023-08-20 manu-draft.qmd
qcreview figure combined in one file using imagemagick convert, added citation to R

The long-term protection and restoration of aquatic resources depends on robust monitoring data; data that require systematic quality control and analysis tools. The *MassWateR* package facilitates quality control, analysis, and data sharing for discrete surface water quality data collected by monitoring programs of various size and technical capacity. The tools were developed to address regional needs for programs in Massachusetts, USA, but the principles and outputs can be applicable to monitoring data collected anywhere.  Users can create quality control reports, perform outlier analyses, and assess trends by season, date, and site for more than 40 parameters. Users can also prepare data for submission to the United States Environmental Protection Agency Water Quality Exchange, thus sharing data to the largest water quality database in the United States. The automated and reproducible workflow offered by *MassWateR* is expected to increase the quantity and quality of publicly available data to support the management of aquatic resources.

@@

The use of R with document generation systems offered through packages like *knitr* [@Xie15] and *rmarkdown* [@Allaire23] can generate QC reports that follow a standard format for review by regulatory agencies.  These tools can also be used to format water quality data for submission to state or national water quality databases, such as the Water Quality Portal (WQP) database maintained by the US Environmental Protection Agency (USEPA) and contributed to via the Water Quality eXchange (WQX).  This database is the largest source of monitoring data in the United States that includes information on hydrologic conditions and chemical, physical, and biological measurements from surface waters.  Further, many environmental resource managers have the need to analyze status and trends in monitoring data and R packages such as *ggplot2* [@Wickham16] and others in the *tidyverse* [@Wickham19] offer useful tools to synthesize and visualize results. Integrating this functionality into a single package is expected to have wide ranging utility for anyone collecting surface water data and is likely to improve the quality and insights obtained from these data.

@@

```{r}
#| echo: true
#| eval: false
qcMWRreview(fset = fsetls, output_dir = getwd())
```
```
#> Report created successfully! File located at /tmp/RtmpUzzrbC/qcreview.docx
```

::: {#fig-qcex layout-ncol=2}

![](images/qcreview1.png)

![](images/qcreview2.png)

The first two of sixteen pages of the quality control report created by `qcMWRreview()` that evaluates the results data relative to data quality objectives.  The first page shows the data quality objectives for accuracy, frequency, and completeness.  The second page shows QC results for frequency and completeness.  Parameters shown in red or marked as 'MISS' failed the data quality objectives.  Users can edit the Word file as needed, e.g., entering the organization name or adding notes.
:::

>>>>

The long-term protection and restoration of aquatic resources depends on robust monitoring data; data that require systematic quality control and analysis tools. The *MassWateR* R package facilitates quality control, analysis, and data sharing for discrete surface water quality data collected by monitoring programs of various size and technical capacity. The tools were developed to address regional needs for programs in Massachusetts, USA, but the principles and outputs can be applicable to monitoring data collected anywhere.  Users can create quality control reports, perform outlier analyses, and assess trends by season, date, and site for more than 40 parameters. Users can also prepare data for submission to the United States Environmental Protection Agency Water Quality Exchange, thus sharing data to the largest water quality database in the United States. The automated and reproducible workflow offered by *MassWateR* is expected to increase the quantity and quality of publicly available data to support the management of aquatic resources.

@@

The use of R [@RCT2023] with document generation systems offered through packages like *knitr* [@Xie15] and *rmarkdown* [@Allaire23] can generate QC reports that follow a standard format for review by regulatory agencies.  These tools can also be used to format water quality data for submission to state or national water quality databases, such as the Water Quality Portal (WQP) database maintained by the US Environmental Protection Agency (USEPA) and contributed to via the Water Quality eXchange (WQX).  This database is the largest source of monitoring data in the United States that includes information on hydrologic conditions and chemical, physical, and biological measurements from surface waters.  Further, many environmental resource managers have the need to analyze status and trends in monitoring data and R packages such as *ggplot2* [@Wickham16] and others in the *tidyverse* [@Wickham19] offer useful tools to synthesize and visualize results. Integrating this functionality into a single package is expected to have wide ranging utility for anyone collecting surface water data and is likely to improve the quality and insights obtained from these data.

@@

```{r}
#| echo: true
#| eval: false
qcMWRreview(fset = fsetls, output_dir = getwd())
```
```
#> Report created successfully! File located at /tmp/RtmpUzzrbC/qcreview.docx
```
![The first two of sixteen pages of the quality control report created by `qcMWRreview()` that evaluates the results data relative to data quality objectives.  The first page shows the data quality objectives for accuracy, frequency, and completeness.  The second page shows QC results for frequency and completeness.  Parameters shown in red or marked as 'MISS' failed the data quality objectives.  Users can edit the Word file as needed, e.g., entering the organization name or adding notes.
](images/qcex.png){#fig-qcex}
